---
title: "p8105_hw2_bz2570"
author: "Boran Zhai"
date: "2025-09-30"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(readr)
```

### Problem 1

Import dataset(nemployment.csv, and snp.csv.)
```{r}
pols_month <- read_csv("./data/pols-month.csv")
unemployment <- read_csv("./data/unemployment.csv")
snp <- read_csv("./data/snp.csv")
```

(1)Clean the data in pols-month.csv. Use separate() to break up the variable mon into integer variables year, month, and day; replace month number with month name; create a president variable taking values gop and dem, and remove prez_dem and prez_gop; and remove the day variable.
```{r}
pols_month <- read_csv("./data/pols-month.csv") |> 
  janitor::clean_names() |>
  separate(mon, into = c("year", "month", "day"), sep = "-") |>
  mutate(
    year = as.integer(year),
    month = as.integer(month),
    day = as.integer(day),
    month = month.abb[month],
    president = ifelse(prez_gop == 1, "gop", 
                       ifelse(prez_dem == 1, "dem",NA))
  ) |>
  select(-prez_dem, -prez_gop, -day)
pols_month
```

(2)Clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.
```{r}

```

